\ttitle

### 2015-08-23 Chapters/CPET_Jaundice.tex
Bug Hunt - CPET Jaundice
- an extra curly }) f%$£ it all up. An hour wasted hunting it down
- This is the actual commit for the error trap
- Not sure why the change didn;'t commit last time
- Maybe because I discarded the change and couldnt stage it again

\chapter{An investigation into the relationship between cardiopulmonary exercise testing, obstructive jaundice and preoperative pathophysiology in patients undergoing major pancreatic surgery.}

@@

\section{Patients and Methods}

\subsection{Patients}
Patients who underwent classical or pylorus-preserving pancreaticoduodenectomy for periampullary lesions (both benign and malignant) between August 2008 and April 2013 and had undergone cardiopulmonary exercise testing as part of their preoperative work-up at the West of Scotland Pancreatic Unit, Glasgow Royal Infirmary, Glasgow were included in the study. 
Established criteria for resectability in patients with malignant disease were used as outlined in previous published work. 
Segmental or wedge resection of the portal vein or superior mesenteric vein was carried out if the lesion was otherwise resectable.

@@

\subsection{Cardiopulmonary exercise test}
Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). 
All patients underwent standard pulmonary function tests and spirometry prior to cardiopulmonary exercise testing. 
A cycle ergometer was used to perform a symptom-limited, incremental work-load test preceded by a 3-minute rest period. 
The test was stopped when patients achieved their maximum exercise tolerance, when significant ischaemic changes occurred on ECG or for other safety reasons. 
Peak oxygen consumption achieved at this stage was defined as $\dot{V}_{O_2}$Peak. 
The $\dot{V}_{O_2}$AT was calculated using the V-slope \parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents \parencite{society_ats/accp_2003} methods.
$\dot{V}_{O_2}$AT less than 10 ml/kg/min was considered to be low based on previous work by us (Chapter \ref{ch_cpet_outcomes}) as well as Ausania and co-workers \parencite{ausania_effects_2012} which has shown increased incidence of complications in patients with $\dot{V}_{O_2}$AT below this threshol. 
Oxygen consumption at peak exercise ($\dot{V}_{O_2}$Peak) was dichotomised using a cut-off of 16 ml/kg/min. 
Cardiopulmonary exercise testing methodology and calculation of $\dot{V}_{O_2}$AT is described in detail in Sections \ref{sec:cpx_method} and \ref{sec:cpx_at_method}.
A mored detailed description of cardiopulmonary exercise testing parameters described in this study are published elsewhere \parencite{balady_clinicians_2010}.

@@

\subsection{Obstructive jaundice vs. preoperative blood tests}
The relationship between obstructive jaundice and preoperative blood tests is shown in Table \ref{table:cpet_oj_bloods}. 
Obstructive jaundice was associated with the presence of a systemic inflammatory response as characterised by raised serum C-reactive protein levels and decrease in serum albumin levels (p$<$0.001). 
The degree of systemic inflammation was proportionate to the severity of jaundice. 
Non-jaundiced patients had a median CRP of 3.6 mg/dl (inter-quartile range 0.3 - 89) and serum albumin of 37 g/dl (IQR 18 - 46) while patients with severe obstructive jaundice had a median CRP of 13 (IQR 1.7 - 51) and a median serum albumin of 25 g/dl (IQR 18 - 33). 
Obstructive jaundice was also associated with electrolyte abnormalities including serum sodium, serum potassium and serum chloride levels. 
Jaundiced patients were more likely to be anaemic (p$<$0.001) with lower haematocrit (p$<$0.001) and lower mean corpuscular volume (p=0.001). 
However, there was no significant difference in the preoperative renal function, prothrombin time and white cell count between jaundiced and non-jaundiced patients.

@@

\subsection{Preoperative factors related to CPET}
Binary logistic regression analysis was undertaken to assess preoperative clinico-pathological patient factors associated with a low $\dot{V}_{O_2}$AT and low $\dot{V}_{O_2}$Peak. 
On univariate analysis, female sex (HR 2.74, 95\% CI 1.30-5.74, p=0.008), body mass index $>$ 25 (HR 3.09, 95\% CI 1.51-6.32, p=0.002), presence of malignancy (HR 3.59, 95\% CI 1.36-9.43, p=0.010), POSSUM Physiology Score $>$ 14 (HR 2.06, 95\% CI 1.02-4.17, p= 0.044), serum bilirubin $>$ 250 $\mu$mol/l (HR 5.66, 95\% CI 1.87-17.16, p=0.002), haemoglobin $<$ 12 g/dl (HR 2.74, 95\% CI 1.30-5.74, p=0.008) and serum C-reactive protein $>$ 10 mg/dl (HR 2.18, 95\% CI 1.06-4.51, p=0.035) were associated with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min. 
Age, cardiovascular and respiratory comorbidity, PBD, mild jaundice and serum albumin were not significantly related to a low $\dot{V}_{O_2}$AT.

The statistically significant independent variables with p$<$0.05 were entered into a backward step-wise regression model with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min as a categorical, binomial, dependent variable. 
Female sex (HR 3.75, 95\% CI 1.57-8.95 p$<$0.005), body mass index $>$ 25 (HR 3.65, 95\% CI 1.61-8.26, p$<$0.005), malignancy (HR 4.02, 95\% CI 1.33-12.16, p$<$0.05) and C-reactive protein $>$ 10 mg/dl (HR 2.98, 95\% CI 1.29-6.86, p$<$0.05) were independently related to low $\dot{V}_{O_2}$AT. 
Obstructive jaundice was not related to low $\dot{V}_{O_2}$AT on multivariate analysis. (Table \ref{table:cpet_oj_at_regression})

@@

%Reasons for not doing PBD
However, several recent studies have reported that PBD is associated with increased incidence of complications and that surgery in the jaundiced patients is safe.
Pitt and coworkers in a prospective randomised trial comparing outcomes in jaundiced patients undergoing surgery with or without PBD reported that PBD was associated with increased cost without any decrease in postoperative complications \parencite{pitt_does_1985}. 
A recent meta-analysis \parencite{sewnath_meta-analysis_2002} pooled data from 5 randomised controlled trials comparing surgery with PBD versus surgery without drainage.
The authors concluded that PBD not only did not improve postoperative complication rates or mortality but resulted in a higher overall complication rate due to the morbidity associated with the procedure itself. 
These findings have been reproduced in a more recent multi-center, randomised trial \parencite{van_der_gaag_preoperative_2010}.
A recent Cochrane Collaboration review of six trials including 520 patients concluded that PBD may be associated with serious adverse events and must not be performed routinely outwith trial settings \parencite{wang_preoperative_2008}.

%CPET vs Jaundice
In the present study, obstructive jaundice was not independently related to $\dot{V}_{O_2}$AT or $\dot{V}_{O_2}$Peak.
There was only a weak negative correlation between serum bilirubin and $\dot{V}_{O_2}$AT in our cohort.
These findings are similar to those of Parker and co-workers who reported that there was only a weak negative correlation between serum bilirubin and  $\dot{V}_{O_2}$Peak (Pearson's $\rho$ -0.21, p=0.02}) and no correlation with $\dot{V}_{O_2}$AT ($\rho$ = -0.15, p=NS) in patients undergoing pancreaticoduodenectomy \parencite{parker_serum_2014}. 
Junejo and co-workers studied oxygen extraction in 9 jaundiced patients during cardiopulmonary exercise test by placing catheters within the femoral vein. 
They reported that oxygen extraction was normal during cardiopulmonary exercise even in the presence of jaundice \parencite{junejo_peripheral_2014}

@@

%Obstructive jaundice and systemic inflammation
However, jaundice was associated with elevated preoperative systemic inflammation as measured by C-reactive protein levels and serum albumin. 
Obstructive jaundice is known to be associated with wide ranging abnormalities of both humoral and innate immune systems \parencite{nehez_compromise_2002}. 
Both cellular immunity and humoral immunity are adversely affected by obstructive jaundice \parencite{padillo_cytokines_2001, scott-conner_pathophysiology_1994}. 
It is unclear if this immune dysfunction before surgery is associated with an abnormal postoperative systemic inflammatory response and this is explored further in Chapter \ref{ch_pre_post_sirs}.
However, the immunological abnormalities associated with obstructive jaundice do not appear to improve after biliary drainage \parencite{kimmings_endotoxin_2000}.

@@

%CPET vs Hemoglobin
The association between low haemoglobin and low $\dot{V}_{O_2}$Peak would appear to be more easily explained since the oxygen carrying capacity of blood depends on haemoglobin concentration.  
Indeed, the linear relationship between oxygen consumption during exercise and haemoglobin concentration has been shown both in disease states \parencite{agostoni_relationship_2010} as well as in healthy volunteers \parencite{dellweg_cardiopulmonary_2008}.
This would suggest that maintaining optimal haemoglobin levels in the perioperative period will improve oxygen delivery and perhaps outcome as well. 
Excessive intra-operative blood loss has been reported to be associated with postoperative complications \parencite{pratt_risk_2008} and this may in part be due impaired oxygen delivery at a time of maximal physiological burden.

>>>>

\chapter{An investigation into the relationship between cardiopulmonary exercise testing, preoperative pathophysiology and obstructive jaundice in patients undergoing pancreaticoduodenectomy.}

@@

\section{Patients and methods}

\subsection{Patients}
Patients who underwent classical or pylorus-preserving pancreaticoduodenectomy for periampullary lesions (both benign and malignant) between August 2008 and April 2013 and had undergone cardiopulmonary exercise testing as part of their preoperative work-up at the West of Scotland Pancreatic Unit, Glasgow Royal Infirmary, Glasgow were included in the study. 
Established criteria for resectability in patients with malignant disease were used as outlined in Section \ref{sec:resectability_criteria}. 
Segmental or wedge resection of the portal vein or superior mesenteric vein was carried out if the lesion was otherwise resectable.

@@

\subsection{Cardiopulmonary exercise test}
Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). 
All patients underwent standard pulmonary function tests and spirometry prior to cardiopulmonary exercise testing. 
A cycle ergometer was used to perform a symptom-limited, incremental work-load test preceded by a 3-minute rest period. 
The test was stopped when patients achieved their maximum exercise tolerance, when significant ischaemic changes occurred on ECG or for other safety reasons. 
Peak oxygen consumption achieved at this stage was defined as $\dot{V}_{O_2}$Peak. 
The $\dot{V}_{O_2}$AT was calculated using the V-slope \parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents \parencite{society_ats/accp_2003} methods.
$\dot{V}_{O_2}$AT less than 10 ml/kg/min was considered to be low based on previous work by us (Chapter \ref{ch_cpet_outcomes}) as well as Ausania and co-workers \parencite{ausania_effects_2012} which has shown increased incidence of complications in patients with $\dot{V}_{O_2}$AT below this threshold. 
Oxygen consumption at peak exercise ($\dot{V}_{O_2}$Peak) was dichotomised using a cut-off of 16 ml/kg/min. 
Cardiopulmonary exercise testing methodology and calculation of $\dot{V}_{O_2}$AT is described in detail in Sections \ref{sec:cpx_method} and \ref{sec:cpx_at_method}.
A more detailed description of cardiopulmonary exercise testing parameters described in this study are published elsewhere \parencite{balady_clinicians_2010}.

@@

\subsection{Obstructive jaundice vs. preoperative blood tests}
The relationship between obstructive jaundice and preoperative blood tests is shown in Table \ref{table:cpet_oj_bloods}. 
Obstructive jaundice was associated with the presence of a systemic inflammatory response as characterised by raised serum C-reactive protein levels and decrease in serum albumin levels (p$<$0.001). 
The degree of systemic inflammation was proportionate to the severity of jaundice. 
Non-jaundiced patients had a median CRP of 3.6 mg/l (inter-quartile range 0.3 - 89) and serum albumin of 37 g/l (IQR 18 - 46) while patients with severe obstructive jaundice had a median CRP of 13 (IQR 1.7 - 51) and a median serum albumin of 25 g/l (IQR 18 - 33). 
Obstructive jaundice was also associated with electrolyte abnormalities including serum sodium, serum potassium and serum chloride levels. 
Jaundiced patients were more likely to be anaemic (p$<$0.001) with lower haematocrit (p$<$0.001) and lower mean corpuscular volume (p=0.001). 
However, there was no significant difference in the preoperative renal function, prothrombin time and white cell count between jaundiced and non-jaundiced patients.

@@

\subsection{Preoperative factors related to CPET}
Binary logistic regression analysis was undertaken to assess preoperative clinico-pathological patient factors associated with a low $\dot{V}_{O_2}$AT and low $\dot{V}_{O_2}$Peak. 
On univariate analysis, female sex (HR 2.74, 95\% CI 1.30-5.74, p=0.008), body mass index $>$ 25 (HR 3.09, 95\% CI 1.51-6.32, p=0.002), presence of malignancy (HR 3.59, 95\% CI 1.36-9.43, p=0.010), POSSUM Physiology Score $>$ 14 (HR 2.06, 95\% CI 1.02-4.17, p= 0.044), serum bilirubin $>$ 250 $\mu$mol/l (HR 5.66, 95\% CI 1.87-17.16, p=0.002), haemoglobin $<$ 12 g/dl (HR 2.74, 95\% CI 1.30-5.74, p=0.008) and serum C-reactive protein $>$ 10 mg/l (HR 2.18, 95\% CI 1.06-4.51, p=0.035) were associated with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min. 
Age, cardiovascular and respiratory comorbidity, PBD, mild jaundice and serum albumin were not significantly related to a low $\dot{V}_{O_2}$AT.

The statistically significant independent variables with p$<$0.05 were entered into a backward step-wise regression model with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min as a categorical, binomial, dependent variable. 
Female sex (HR 3.75, 95\% CI 1.57-8.95 p$<$0.005), body mass index $>$ 25 (HR 3.65, 95\% CI 1.61-8.26, p$<$0.005), malignancy (HR 4.02, 95\% CI 1.33-12.16, p$<$0.05) and C-reactive protein $>$ 10 mg/l (HR 2.98, 95\% CI 1.29-6.86, p$<$0.05) were independently related to low $\dot{V}_{O_2}$AT. 
Obstructive jaundice was not related to low $\dot{V}_{O_2}$AT on multivariate analysis. (Table \ref{table:cpet_oj_at_regression})

@@

%Reasons for not doing PBD
However, several recent studies have reported that PBD is associated with increased incidence of complications and that surgery in the jaundiced patients is safe.
Pitt and co-workers in a prospective randomised trial comparing outcomes in jaundiced patients undergoing surgery with or without PBD reported that PBD was associated with increased cost without any decrease in postoperative complications \parencite{pitt_does_1985}. 
A recent meta-analysis \parencite{sewnath_meta-analysis_2002} pooled data from 5 randomised controlled trials comparing surgery with PBD versus surgery without drainage.
The authors concluded that PBD not only did not improve postoperative complication rates or mortality but resulted in a higher overall complication rate due to the morbidity associated with the procedure itself. 
These findings have been reproduced in a more recent multicenter, randomised trial \parencite{van_der_gaag_preoperative_2010}.
A recent Cochrane Collaboration review of six trials including 520 patients concluded that PBD may be associated with serious adverse events and must not be performed routinely outside trial settings \parencite{wang_preoperative_2008}.

%CPET vs Jaundice
In the present study, obstructive jaundice was not independently related to $\dot{V}_{O_2}$AT or $\dot{V}_{O_2}$Peak.
There was only a weak negative correlation between serum bilirubin and $\dot{V}_{O_2}$AT in our cohort.
These findings are similar to those of Parker and co-workers who reported that there was only a weak negative correlation between serum bilirubin and  $\dot{V}_{O_2}$Peak (Pearson's $\rho$ -0.21, p=0.02} and no correlation with $\dot{V}_{O_2}$AT ($\rho$ = -0.15, p=NS) in patients undergoing pancreaticoduodenectomy \parencite{parker_serum_2014}. 
Junejo and co-workers studied oxygen extraction in 9 jaundiced patients during cardiopulmonary exercise test by placing catheters within the femoral vein. 
They reported that oxygen extraction was normal during cardiopulmonary exercise even in the presence of jaundice \parencite{junejo_peripheral_2014}

@@

%Obstructive jaundice and systemic inflammation
However, jaundice was associated with elevated preoperative systemic inflammation as measured by C-reactive protein levels and serum albumin. 
Obstructive jaundice is known to be associated with wide ranging abnormalities of both humoral and innate immune systems \parencite{nehez_compromise_2002, padillo_cytokines_2001, scott-conner_pathophysiology_1994}. 
It is unclear if this immune dysfunction before surgery is associated with an abnormal postoperative systemic inflammatory response and this is explored further in Chapter \ref{ch_pre_post_sirs}.
However, the immunological abnormalities associated with obstructive jaundice do not appear to improve after biliary drainage \parencite{kimmings_endotoxin_2000}.

@@

%CPET vs Haemoglobin
The association between low haemoglobin and low $\dot{V}_{O_2}$Peak would appear to be more easily explained since the oxygen carrying capacity of blood depends on haemoglobin concentration.  
Indeed, the linear relationship between oxygen consumption during exercise and haemoglobin concentration has been shown both in disease states \parencite{agostoni_relationship_2010} as well as in healthy volunteers \parencite{dellweg_cardiopulmonary_2008}.
This would suggest that maintaining optimal haemoglobin levels in the perioperative period will improve oxygen delivery and perhaps outcome as well. 
Excessive intra-operative blood loss has been reported to be associated with postoperative complications \parencite{pratt_risk_2008} and this may in part be due impaired oxygen delivery at a time of maximal physiological burden.
